---
title: "analysis"
format: html
editor_options: 
  chunk_output_type: console
---

# Load packages and scripts

```{r}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(here, gert, tidyverse, assertthat, glue, readxl, sf, janitor, 
               sdmTMB, reticulate, cropgrowdays, pipebind, cmdstanr, loo, 
               posterior, tidybayes, MetBrewer)
options(mc.cores = 8)

# if (dir.exists("aru-occ")) {
#   git_pull(repo = "aru-occ")
# } else {
#   git_clone("https://github.com/mhollanders/aru-occ")
# }
source(here("aru-occ/R/aru-functions.R"))
source(here("aru-occ/R/util.R"))
source(here("R/my-theme.R"))
theme_set(my_theme())
```

# Read and process data

changelog
- changed some incorrect dates missing slashes etc.
- removed dates with "." 
- remove corrected_date entries with "-" and text; must be a date
- number of individuals has to be a number; change to minimum of range if given
- some records are e.g. "rabbit and hare". these should be split ( I haven't)
- row 93 in 2020_detections was 2010; row 95 was 2021
- "wasn't active" changed to one date before last download (so that removed days are early death + days(1))


## Deployments and detections

```{r}
# sheets in data
loc <- here("scratch/south-coast/Cocrackerup_Peniup_RedMoort_cameradata.xlsx")
sheets <- excel_sheets(loc)

# site locations
locs_raw <- map(list.files(here("scratch/south-coast"), pattern = "\\.shp", full.names = T),
                ~ read_sf(.) |> clean_names()) |>
  bind_rows()

locs <- locs_raw |>
  bind_cols(st_coordinates(locs_raw)) |>
  select(site = name, longitude = X, latitude = Y) |>
  add_utm_columns(units = "km")

# deployment stuff
sites_raw <- read_excel(loc, sheet = sheets[[2]]) |>
  clean_names() |>
  rename_with(~ paste0("early_death_", seq_along(.x)),
              .cols = starts_with("early_death")) |>
  rename_with(~ paste0("download_", seq_along(.x)),
              .cols = starts_with("download")) |>
  mutate(across(where(is.POSIXct), ymd), site = waypoint)

sites <- sites_raw |>
  mutate(start = ymd(date_set_up),
         end = ymd(sites_raw[[ncol(sites_raw)]]),
         .after = waypoint) |>
  left_join(locs, by = "site") |>
  mutate(site = factor(site, levels = str_sort(unique(site), numeric = T)))

failures <- sites |>
  pivot_longer(cols = matches("^(early_death_|download_)"),
               names_to = c(".value", "period"),
               names_pattern = "(early_death|download)_(\\d+)") |>
  drop_na(early_death)

# detections
dets_raw <- read_excel(loc, sheet = sheets[[1]]) |>
  clean_names()

dets <- dets_raw |>
  mutate(date = ymd(date),
         time = hms::as_hms(time),
         timestamp = ymd_hms(str_c(date, time)),
         site = factor(location, levels = levels(sites$site)),
         species = case_when(species == "Black-gloved wallaby" ~ "Brush wallaby",
                             .default = species)) |>
  filter(!str_detect(spp_category, "Bird|Reptile|unidentified|no ID|Unidentified"),
         !str_detect(species, "no ID|eagle|Honey")) |>
  mutate(species = factor(str_to_title(species))) |>
  arrange(species, timestamp) |>
  select(site, timestamp, species, no_individuals) |> 
  drop_na()

# labels
site_lvl <- levels(sites$site)
species_lvl <- levels(dets$species)
```

## Site predictors

```{r}
# setup; this was a pain 
Sys.setenv("RETICULATE_PYTHON" = "/Users/matt.hollanders/.virtualenvs/rgee/bin/python")
Sys.setenv("EARTHENGINE_GCLOUD" = "/Users/matt.hollanders/google-cloud-sdk/bin/")
use_virtualenv("rgee", required = TRUE)
library(rgee)
ee <- import("ee")
ee$Authenticate()
ee$Initialize(project = "heroic-rain-367400")
ee_dir <- "~/.config/earthengine"
if (!dir.exists(ee_dir)) {
  dir.create(ee_dir, recursive = TRUE, showWarnings = FALSE)
  write.table(data.frame(user = "matthijshollanders", 
                         drive_cre = NA, 
                         gcs_cre = NA),
              file = path.expand(str_c(ee_dir, "/rgee_sessioninfo.txt")),
              row.names = F)
}

# prepare sites as EE object
deployments_to_fc <- function(deployments, site = site, 
                              longitude = "longitude", latitude = "latitude", 
                              crs = 4326) {
  deployments |> 
    st_as_sf(coords = c(longitude, latitude), crs = crs) |> 
    select({{ site }}) |>
    sf_as_ee()
}

# terrain function
extract_terrain <- function(deployments, site = site) {
  dem <- ee$Image("USGS/SRTMGL1_003")
  info <- dem$rename("elevation")$
    addBands(ee$Terrain$slope(dem)$rename("slope"))$
    addBands(ee$Terrain$aspect(dem)$rename("aspect"))$
    reduceRegions(collection = deployments_to_fc(deployments),
                  reducer = ee$Reducer$mean(),
                  scale = 30)$
    getInfo()
  site_lvl <- pull(deployments, {{ site }}) |> levels()
  info$features |> 
    map(~as_tibble(.$properties)) |> 
    list_rbind() |> 
    relocate({{ site }}, .before = 1) |> 
    mutate({{ site }} := factor({{ site }}, levels = site_lvl))
}
ter <- extract_terrain(sites) |> 
  mutate(aspect_radian = aspect * pi / 180,
         northness = cos(aspect_radian),
         eastness = sin(aspect_radian)) |> 
  select(site, elevation, slope, northness, eastness)
ter_scaled <- ter |> 
  mutate(aspect_radian = aspect * pi / 180,
         northness = cos(aspect_radian),
         eastness = sin(aspect_radian)) |> 
  select(site, elevation, slope, northness, eastness) |> 
  mutate(across(-site, ~scale(.)[, 1]))
```


## Survey predictors

### Meteorological

```{r}
# extract SILO data
silo_loc <- here("data/silo.rds")
if (file.exists(silo_loc)) {
  silo <- read_rds(silo_loc)
} else {
  silo <- get_multi_silodata(latitude = sites$latitude,
                             longitude = sites$longitude,
                             Sitename = sites$site,
                             email = "matthijs.hollanders@gmail.com",
                             START = min(sites$start) |> str_remove_all("-"),
                             FINISH = max(sites$end) |> str_remove_all("-"))
  write_rds(silo, silo_loc)
}

# select some and convert to long format
silo_long <- silo |> 
  mutate(sqrt_rain = sqrt(rain)) |> 
  select(site = Sitename, date = date_met, maxt, sqrt_rain, vp) |> 
  pivot_longer(-c(site, date))

# summarise means and SDs using date-level averages
silo_stats <- silo_long |> 
  summarise(average_value = mean(value), 
            .by = c(date, name)) |> 
  summarise(mean = mean(average_value),
            sd = sd(average_value),
            .by = name)

silo

# scale, but not center rain
silo_scaled <- silo_long |> 
  left_join(silo_stats) |> 
  mutate(value_scaled = (value - if_else(name == "sqrt_rain", 0, mean)) / sd) |> 
  select(site, date, name, value_scaled)
```

### Vegetation indices

```{r}
# function extract NDVI and EVI
extract_vegetation_index <- function(deployments, site = site, start = start, 
                                     end = end) {
  # get start and end dates and site levels
  date_start <- pull(deployments, {{ start }}) |> min()
  date_end <- pull(deployments, {{ end }}) |> max()
  site_lvl <- pull(deployments, {{ site }}) |> levels()
  
  # filter
  ic <- ee$ImageCollection("MODIS/061/MOD13A1")$
    filterDate(as.character(date_start - days(16)), 
               as.character(date_end + days(16)))$
    select(c("NDVI", "EVI"))$
    map(\(img) img$multiply(0.0001))
  
  # extract
  info <- ic$toBands()$
    reduceRegions(collection = deployments_to_fc(deployments),
                  reducer = ee$Reducer$mean())$
    getInfo()
    
  # out
  info$features |>
    map(~as_tibble(.$properties)) |>
    list_rbind() |> 
    mutate({{ site }} := factor({{ site }}, levels = site_lvl)) |> 
    pivot_longer(-{{ site }}, names_to = "band") |> 
    separate_wider_regex(band, 
                         patterns = c(date = ".*", "_", index = "NDVI|EVI")) |> 
    mutate(date = ymd(date))
}
vi <- extract_vegetation_index(sites)

# scale using date-level averages
vi_stats <- vi |> 
  summarise(average_value = mean(value), .by = c(index, date)) |> 
  summarise(mean = mean(average_value),
            sd = sd(average_value),
            .by = index)
  
vi_scaled <- vi |> 
  left_join(vi_stats, by = "index") |> 
  mutate(value_scaled = (value - mean) / sd) |> 
  select(site, date, index, value_scaled)
```

### Combine

```{r}
survey_predictors <- silo_scaled |>   
  pivot_wider(names_from = name, values_from = value_scaled) |> 
  left_join(
    vi_scaled |> 
      pivot_wider(names_from = index, values_from = value_scaled) |> 
      complete(site, date = seq.Date(min(date), max(date))) |> 
      fill(EVI, NDVI, .direction = "down"),
    by = c("site", "date")
  ) |> 
  select(-NDVI) |> 
  mutate(camera_fct = 
           case_when((str_detect(site, "CK") & date >= ymd("2023-08-17") |
                        str_detect(site, "PN") & date >= ymd("2023-08-20") | 
                        str_detect(site, "RM") & date >= ymd("2023-10-11")) ~ "Swift",
                     .default = "Scoutguard") |> 
           factor(levels = c("Scoutguard", "Swift")),
         camera = as.integer(camera_fct) - 1.5) |> 
  select(-camera_fct)

date <- sym("date")
reference_date <- min(sites$start) 

survey_summary = list("sqrt_rain" = sum)
summary_fns <- setNames(
      lapply(X3_lvl, \(col) {
        if (!is.null(survey_summary) && col %in% names(survey_summary)) {
          match.fun(survey_summary[[col]])
        } else {
          mean
        }
      }),
      X3_lvl
    )
X3_lvl <- colnames(survey_predictors[, -c(1:2)])
P <- list()
P[3] <- length(X3_lvl)

X3 <- survey_predictors |>
      dplyr::filter({{ date }} >= reference_date) |>
      dplyr::mutate(date = aggregate_by_days({{ date }}, reference_date, days)) |>
      dplyr::summarise(dplyr::across(dplyr::where(is.numeric), 
                                     ~summary_fns[[dplyr::cur_column()]](., na.rm = TRUE)),
                       .by = c({{ deploymentID }}, date)) |>
      tidyr::pivot_longer(-c({{ deploymentID }}, date), names_to = "p") |>
      tidyr::complete({{ deploymentID }} := factor(site_lvl, site_lvl),
                      date := lubridate::ymd(surveys),
                      p := factor(X3_lvl, X3_lvl),
                      fill = list(value = 0)) |>
      dplyr::arrange(date, p, {{ deploymentID }}) |>
      dplyr::pull(value) |> 
      array(c(I, P[3], J), dimnames = list(site_lvl, X3_lvl, surveys))
    if (scale) {
      survey_means <- apply(X3, c(2, 3), mean)
      X3_means <- apply(survey_means, 1, mean)
      X3_sds <- apply(survey_means, 1, sd)
      X3 <- sweep(sweep(X3, 2, X3_means, "-"), 2, X3_sds, "/")
    }
dim(X3)
    
    survey_predictors
    survey_summary <- list("rain" = sum)


first_last <- survey_predictors |>
  dplyr::summarise(first = min({{ date }}), last = max({{ date }}),
                   .by = {{ deploymentID }}) |> 
  left_join(
    deployments, by = join_by({{ deploymentID }})
  )
    if (first_last$first > reference_date ||
        first_last$last < max(dplyr::pull(deployments, {{ deploymentEnd }}))) {
      cli::cli_abort(
        "{.arg survey_predictors} must cover the full deployment period."
      )
    }
```

# Stan

## Prepare data

```{r}
# data prep for Stan
dh <- prep_data(sites, dets, count = no_individuals, 
                projected_X = X, projected_Y = Y,
                failures = failures, failure_start = early_death,
                failure_end = download, days = 7, minutes = 30,
                # occupancy_site_predictors = ter_scaled,
                detection_site_predictors = ter_scaled,
                survey_predictors = survey_predictors
                ) |> 
  append_defaults(D = 50, grainsize = 0, OD = 0) |> 
  glimpse()
apply(dh$y, 3, sum)


# transform survey predictors
dh$X3[, "rain", ] <- sqrt(dh$X3[, "rain", ])
for (p in 1:dh$P[3]) {
  dh$X3[, p, ] <- (dh$X3[, p, ] - mean(colMeans(dh$X3[, p, ]))) / sd(colMeans(dh$X3[, p, ]))
}
```

## Fit models

```{r}
mod <- cmdstan_model(here("aru-occ/stan/aru-occ-ms.stan"), 
                     cpp_options = list(stan_threads = T))
register_default_progress_handler()
mod_lvl <- c("Poisson", "Poisson OLRE", "Negative Binomial", "Negative Binomial w/o kappa")
configs <- expand_grid(OD = 0:2) |> 
  mutate(config = row_number(), .before = 1)
fits <- map(configs$config, ~{
  dh$OD <- configs$OD[.]
  mod$sample(dh, chains = 2, threads_per_chain = 4, 
             iter_warmup = 200, iter_sampling = 200, show_exceptions = T,
             max_treedepth = 8, show_progress_bar = T, 
             init = mod$pathfinder(dh, init = 0.1, sig_figs = 14,
                                   num_threads = 2, num_paths = 2,
                                   single_path_draws = 100,
                                   max_lbfgs_iters = 200,
                                   psis_resample = F))
})
hypers <- c(str_c(c("psi", "mu", "beta"), "_bar"), "W", "theta")
fits[[2]]$draws(hypers) |> mcmc_trace()

map(fits, ~gather_rvars(., beta[d, s, p], beta_unc[s, p]) |> 
      filter(d == 2 | is.na(d))) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(xdist = .value, y = factor(s, labels = species_lvl))) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  facet_grid(factor(p, labels = rownames(dh$X2)) ~ .variable) +
  stat_pointinterval(aes(colour = factor(mod)),
                     position = position_dodge(width = 0.3)) + 
  scale_x_continuous(limits = c(-2, 2))

fit2 |> 
  gather_rvars(beta[d, s, p], beta_unc[s, p]) |> 
  filter(d == 2 | is.na(d)) |> 
  ggplot(aes(xdist = .value, y = factor(s, labels = species_lvl))) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  facet_wrap(~ factor(p, labels = rownames(dh$X2))) +
  stat_pointinterval(aes(colour = .variable),
                     position = position_dodge(width = 0.3)) + 
  scale_x_continuous(limits = c(-2, 2))

fit |> 
  gather_rvars(iota[s, i], iota_unc[s, i]) |> 
  ggplot(aes(xdist = .value, y = factor(i, labels = site_lvl) |> fct_rev())) + 
  facet_wrap(~ factor(s, labels = species_lvl), nrow = 1, scales = "free_x") + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") +
  stat_pointinterval(aes(colour = .variable),
                     .width = 0.95, size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.5)) + 
  scale_x_continuous(limits = c(-2, 2))

map(list(fit, fit2), ~gather_rvars(., beta[d, s, p]) |> 
      filter(d == 2 | is.na(d))) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(xdist = .value, y = factor(s, labels = species_lvl))) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  facet_wrap(~ factor(p, labels = rownames(dh$X2))) +
  stat_pointinterval(aes(colour = factor(mod)),
                     position = position_dodge(width = 0.3))
  

fit |> 
  gather_rvars(beta_bar[d, p], beta_bar_unc[p]) |> 
  filter(d == 1 | is.na(d)) |> 
  ggplot(aes(xdist = .value, y = factor(p, labels = rownames(dh$X2)))) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(colour = .variable),
                     position = position_dodge(width = 0.3))

fits[[2]] |> 
  gather_rvars(gamma[s, p], gamma_unc[s, p]) |> 
  ggplot(aes(xdist = .value, y = factor(s, labels = species_lvl))) + 
  facet_wrap(~ factor(p, labels = colnames(dh$X3))) +
  stat_pointinterval(aes(colour = .variable),
                     position = position_dodge(width = 0.3))
fit |> 
  gather_rvars(iota[s, i], iota_unc[s, i]) |> 
  ggplot(aes(xdist = .value, y = factor(i, labels = site_lvl) |> fct_rev())) + 
  facet_wrap(~ factor(s, labels = species_lvl)) +
  stat_pointinterval(aes(colour = .variable),
                     position = position_dodge(width = 0.3))


map(list(fits[[2]], fits2[[2]]), ~gather_rvars(., gamma[s, p], gamma_unc[s, p])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(xdist = .value, y = factor(s))) + 
  facet_grid(.variable ~ p) + 
  stat_pointinterval(aes(colour = factor(mod)),
                     position = "dodge")

```

## Goodness-of-fit

```{r}
# loo
loos <- map(fits, ~.$loo("log_lik2"))
loo_compare(loos)
map(fits, ~spread_rvars(., log_lik2[s, i])) |>
  list_rbind(names_to = "config") |> 
  ggplot(aes(xdist = log_lik2, y = factor(i, labels = site_lvl) |> fct_rev())) +
  facet_wrap(~ factor(s, labels = species_lvl), nrow = 1, scales = "free_x") + 
  stat_pointinterval(aes(colour = factor(config)), 
                     point_interval = median_qi, .width = 0.95,
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3))
map(loos, ~as_tibble(.$pointwise) |> 
      mutate(i = row_number())) |> 
  list_rbind(names_to = "config") |> 
  ggplot(aes(influence_pareto_k, i)) + 
  geom_point(aes(colour = config))
loos[[1]]$pointwise |> as_tibble()

# PPC
glimpse(dh)
pp_check(apply(dh$y, c(1, 3), sum) |> c(),
         yrep = fits[[3]]$draws("Qrep", format = "draws_matrix"),
         group = rep(species_lvl, each = dh$I),
         fun = ppc_rootogram_grouped,
         style = "suspended") + 
  scale_x_continuous(limits = c(0, 50))
?ppc_rootogram

```

# Plots

## Intercepts

```{r}
map(fits, ~spread_rvars(., alpha[s, d])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(xdist = if_else(d == 1, inv_logit(alpha), alpha), 
             y = factor(s, labels = species_lvl) |> fct_rev(),
             colour = factor(mod, labels = mod_lvl))) + 
  facet_wrap(~ factor(d, labels = c("Occupancy Probability", "Log Detection Rate")), 
             scale = "free_x", nrow = 1) + 
  stat_pointinterval(point_interval = median_hdci, .width = 0.95,
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.2)) + 
  labs(x = "Estimate", y = "Species", colour = "Observation Model")
```


## Realised occupancy

```{r}
map(fits, ~spread_rvars(., z[i, s]) |> 
      summarise(Z = rvar_sum(z), .by = s)) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(xdist = Z / dh$I, 
             y = factor(s, labels = species_lvl) |> fct_rev(),
             colour = factor(mod, labels = mod_lvl))) + 
  stat_pointinterval(point_interval = median_hdci, .width = 0.95,
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.2)) + 
  scale_x_continuous(breaks = seq(0.2, 0.8, 0.2), limits = c(0, 1), expand = c(0, 0)) + 
  labs(x = "Proportion of sites occupied", y = "Species", colour = "Observation Model")

```

## Variance partitions

```{r}
v_lbl <- c("Species intercept", 
           str_c(rownames(dh$X1), " Global"),
           str_c(rownames(dh$X1), " Species"),
           str_c(colnames(dh$X3), " Global"),
           str_c(colnames(dh$X3), " Species"),
           str_c("Site Effect ", c("Global", species_lvl)),
           str_c("Survey Effect ", c("Global", species_lvl))
           # str_c("Residual ", c("Global", species_lvl))
           )
map(fits, ~spread_rvars(., W_phi[v, d]) |> 
      filter(mean(W_phi) > 0)) |> 
  list_rbind(names_to = "mod")
  
fit |> spread_rvars(W_phi[v, d]) |> 
  ggplot(aes(xdist = W_phi, 
             y = factor(v, labels = v_lbl) |> fct_rev(),
             # colour = factor(mod, labels = mod_lvl)
             )) + 
  facet_wrap(~ factor(d, labels = c("Occupancy", "Detection")), 
             scales = "free_x", nrow = 1) + 
  stat_pointinterval(point_interval = median_hdci, .width = 0.95,
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.5)) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_met_d("Kandinsky") + 
  labs(x = "Partition", y = "Component", colour = "Observation Model")
ggsave(here("manuscript/figs/fig-W_phi.png"), width = 8, height = 12, dpi = 600)

```


## Coefficients

```{r}
# site effects
map(fits, ~spread_rvars(., beta_bar[d, p], beta[d, s, p])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(y = factor(s, labels = species_lvl) |> fct_rev())) + 
  facet_grid(factor(p, labels = rownames(dh$X1)) ~ 
               factor(d, labels = c("Occupancy", "Detection")),
             scales = "free_x") + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = beta, 
                         colour = factor(mod)), #  labels = mod_lvl)), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Predictor", colour = "Observation Model")
map(fits, ~spread_rvars(., beta_unc[s, p])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(y = factor(s, labels = species_lvl) |> fct_rev())) + 
  facet_wrap(~ factor(p, labels = rownames(dh$X1)),
             scales = "free_x", nrow = 1) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = beta_unc, 
                         colour = factor(mod)), #  labels = mod_lvl)), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Predictor", colour = "Observation Model")

map(fits, ~spread_rvars(., beta_bar[d, p], beta[d, s, p]) |> 
      filter(d == 2)) |> 
  list_rbind(names_to = "config") |> 
  left_join(configs) |> 
  ggplot(aes(y = factor(s, labels = species_lvl) |> fct_rev())) + 
  facet_grid(factor(p, labels = rownames(dh$X1)) ~ 
               factor(OD, labels = mod_lvl),
             scales = "free_x") + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = beta, 
                         colour = factor(RSR)), #  labels = mod_lvl)), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Predictor", colour = "Observation Model")

# survey effects
map(fits, ~spread_rvars(., gamma_bar[p], gamma[s, p], gamma_unc[s, p])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(y = factor(p, labels = colnames(dh$X3)))) + 
  facet_wrap(~ factor(s, labels = species_lvl), nrow = 2) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = gamma_unc, 
                         colour = factor(mod)), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Predictor", colour = "Observation Model")

map(fits[3:4], ~spread_rvars(., gamma_bar[p], gamma[s, p])) |> 
  list_rbind(names_to = "mod") |> 
  ggplot(aes(y = factor(s, labels = species_lvl) |> fct_rev())) + 
  facet_wrap(~ factor(p, labels = colnames(dh$X3)), 
             nrow = 1, scales = "free_x") + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = gamma, 
                         colour = factor(mod, labels = c("Negative Binomial", "Negative Binomial w/o kappa"))), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  scale_colour_met_d("Kandinsky") + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Species", colour = "Observation Model")
ggsave(here("manuscript/figs/fig-gamma2.png"), width = 10, height = 7, dpi = 600)

map(fits, ~spread_rvars(., gamma_bar[p], gamma[s, p])) |> 
  list_rbind(names_to = "config") |> 
  left_join(configs) |> 
  ggplot(aes(y = factor(s, labels = species_lvl) |> fct_rev())) + 
  facet_grid(factor(p, labels = colnames(dh$X3)) ~ factor(OD, labels = mod_lvl)) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(xdist = gamma, colour = factor(RSR)), 
                     point_interval = median_hdci, .width = 0.95, 
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.3)) + 
  scale_colour_met_d("Kandinsky") + 
  # stat_pointinterval(aes(xdist = gamma_bar), 
  #                    point_interval = median_hdci, .width = 0.95, 
  #                    size = 0.5, linewidth = 0.5,
  #                    position = position_nudge(y = -0.2), colour = "red4")
  labs(x = "Estimate", y = "Species", colour = "Observation Model")
```

## Correlations

```{r}
fits[[4]] |> 
  spread_rvars(O[g, s, ss]) |> 
  mutate(s = factor(s, labels = species_lvl),
         ss = factor(ss, labels = species_lvl)) |> 
  filter(median(O) != 1) |> 
  ggplot(aes(xdist = O, y = fct_rev(s))) + 
  facet_grid(ss ~ factor(g, labels = c("Survey predictors", "Site", "Survey", "Residual")[1:5])) + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(point_interval = median_hdci, .width = 0.95,
                     size = 0.5, linewidth = 0.5) +
  labs(x = "Correlation", y = "Species")
ggsave(here("manuscript/figs/fig-O.png"), width = 10, height = 15, dpi = 600)
```

## Site effects

```{r}
map(fits, ~spread_rvars(., iota[s, i])) |> 
  list_rbind(names_to = "config") |> 
  left_join(configs) |> 
  ggplot(aes(xdist = iota, y = factor(i, labels = site_lvl) |> fct_rev())) + 
  facet_grid(factor(OD, labels = mod_lvl) ~ factor(s, labels = species_lvl),
             scales = "free_x") + 
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red4") + 
  stat_pointinterval(aes(colour = factor(RSR)),
                     point_interval = median_qi, .width = 0.95,
                     size = 0.5, linewidth = 0.5,
                     position = position_dodge(width = 0.5)) + 
  scale_colour_met_d("Egypt") +
  labs(x = "iota", y = "Site", colour = "RSR")
```


## Survey effects

```{r}
fits[[4]] |> 
  spread_rvars(alpha[s, d], kappa[s, j], kappa_bar[j]) |> 
  filter(d == 2) |> 
  ggplot(aes(factor(j, labels = colnames(dh$Delta)) |> ymd())) + 
  facet_wrap(~ factor(s, labels = species_lvl), scales = "free_y", ncol = 1) + 
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red4") +
  # stat_lineribbon(aes(ydist = exp(alpha + kappa_bar)), fill = "red4",
  #                point_interval = median_qi, linewidth = 0.5, alpha = 0.2) +
  # geom_point(aes(ymd(survey), count),
  #            data = reshape2::melt(dh$y,
  #                                  value.name = "y", 
  #                                  varnames = c("site", "survey", "species")) |> 
  #              as_tibble() |> 
  #              left_join(reshape2::melt(dh$Delta,
  #                                       value.name = "Delta", 
  #                                       varnames = c("site", "survey")),
  #                        by = c("site", "survey")) |> 
  #              filter(Delta > 0) |> 
  #              summarise(count = mean(y / Delta), .by = c(survey, species)),
  #            alpha = 0.2) + 
  stat_lineribbon(aes(ydist = exp(alpha + kappa)),
                  point_interval = median_qi, linewidth = 0.5, alpha = 0.5, 
                  show.legend = F) +
  scale_fill_met_d("VanGogh3") + 
  scale_x_date(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0), n.breaks = 4) + 
  # scale_y_continuous(breaks = seq(0.5, 2, 0.5), limits = c(0, 2.25), expand = c(0, 0)) + 
  labs(x = "Date", y = "Weekly Detection Rate")
ggsave(here("manuscript/figs/fig-kappa.png"), width = 7, height = 10, dpi = 600)
```



```{r}

gc()

# Packages ======
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)
library(ggspatial)
library(cowplot)
library(dplyr)
library(glue)
library(httr)
library(utils)
library(terra)
library(tmap)
library(lubridate)
library(raster)
library(gstat)



# High res map of Australia
# Define the URL for the GDA2020 shapefile
url <- "https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/access-and-downloads/digital-boundary-files/aus_2021_aust_shp_gda2020.zip"
# First accessed: 26th May 2025
# Australian Bureau of Statistics - Digital Boundary Files
# CRS: GDA2020 ESG7844 

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Australia boundary")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "aus_2021_aust_shp_gda2020.zip")

# Download the file
download.file(url, destfile = dest_file, mode = "wb")

# Unzip the file
unzip(dest_file, exdir = dest_dir)

# Optional: Remove the zip file after extraction
file.remove(dest_file)

# Define the path to the shapefile (adjust the filename as needed)
shapefile_path <- file.path(dest_dir, "AUS_2021_AUST_GDA2020.shp")
shapefile_path

# Read the shapefile
aus_boundary <- st_read(shapefile_path)

# Check the CRS
st_crs(aus_boundary)

# Structure
str(aus_boundary)

# Plot
plot(st_geometry(aus_boundary), main = "Australia")



# Generate study area extent from movement data ================================
# Will be used to crop and mask spatial layers
# Define extent in which to generate locations
crop_box <- st_as_sfc(st_bbox(c(
  xmin = 118.899755,
  xmax = 119.340872,
  ymin = -35.549294,
  ymax = -34.326211
), crs = 7844))

# Crop Australian land to the extent
study_area <- st_intersection(aus_boundary, crop_box)

# Plot
plot(st_geometry(study_area), col = "lightgreen", main = "Cropped Australia Land Area")
st_crs(study_area)

# Create bounding box - define extent
study_bbox <- st_as_sfc(st_bbox(study_area))
study_bbox
plot(study_bbox)


# Prepare crop and mask area for other spatial layers
# Set seed for reproducibility
set.seed(42)

# Generate random sample points within the polygon
n_points <- 1000
gps_sf <- st_sample(study_area, size = n_points, type = "random")

# Convert to an sf object (st_sample returns geometry only)
gps_sf <- st_sf(geometry = gps_sf)

# Plot
plot(st_geometry(study_area), main = "Sample Points in Polygon")
plot(gps_sf, add = TRUE, col = "blue", pch = 16)

# Convert buffer distance (1000 m) to degrees
lat <- -34.4
meters_per_degree_lon <- 111000 * cos(lat * pi / 180)
meters_per_degree_lat <- 111000
buffer_lon_deg <- 1000 / meters_per_degree_lon
buffer_lat_deg <- 1000 / meters_per_degree_lat

# Rename and extract bbox
bbox_points <- st_bbox(gps_sf)

# Use the correct bbox object
bbox_expanded <- bbox_points  # this is your named numeric vector

# Apply the buffer in degrees
bbox_expanded["xmin"] <- bbox_points["xmin"] - buffer_lon_deg
bbox_expanded["xmax"] <- bbox_points["xmax"] + buffer_lon_deg
bbox_expanded["ymin"] <- bbox_points["ymin"] - buffer_lat_deg
bbox_expanded["ymax"] <- bbox_points["ymax"] + buffer_lat_deg

# # Buffer the polygon by 1000 meters (or 500 m if you prefer) - when working with projected coordinates
# buffer_distance <- 1000  # in meters (since CRS is UTM)
# bbox_buffered <- st_buffer(bbox_polygon, dist = buffer_distance)
# bbox_buffered
# 
# # Create a bounding box from the buffered box to get square corners
# bbox <- st_bbox(bbox_buffered)
# 
# # Convert bbox to an sf polygon (sfc object)
# bbox_polygon <- st_as_sfc(bbox)
# bbox_polygon

# Convert to polygon
bbox_polygon <- st_as_sfc(bbox_expanded)

# Optional: plot to confirm
plot(st_geometry(study_area), col = "lightgreen", main = "Buffered Bounding Box")
plot(bbox_polygon, add = TRUE, border = "red", lwd = 2)

# Use bbox_polygon to crop spatial layers
# Crop Australian land to Bremer Bay study area extent
st_crs(aus_boundary)==st_crs(bbox_polygon) # Check geometries match
bb_study_area <- st_intersection(aus_boundary, bbox_polygon)
plot(bb_study_area["AUS_CODE21"])

# Reproject
bb_study_area_7850 <- st_transform(bb_study_area, crs = 7850) # UTM zone 50 projected CRS

# Transform points to match the study area CRS
gps_sf_transformed <- st_transform(gps_sf, st_crs(bb_study_area_7850))

# Plot to check point buffer has worked
plot(st_geometry(bb_study_area_7850), main = "Australia")
plot(gps_sf_transformed, add = TRUE, col = "blue", pch = 16)


# Save bounding box of study area
# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/Study area bounding box")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "Bremer Bay study area bounding box.shp")

# Define the path to the shapefile (adjust the filename as needed)
shapefile_path <- file.path(dest_dir, "Bremer Bay study area bounding box.shp")
shapefile_path

# Save study area bounding box 
st_write(bbox_polygon,
         dsn = shapefile_path,
         delete_layer = TRUE)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Study area bounding box"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".shp", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
file_path
bbox_polygon <- st_read(file_path)
plot(bbox_polygon)
class(bbox_polygon)

# Now convert to terra
bbox_polygon_spat <- vect(bbox_polygon)
class(bbox_polygon_spat)
str(bbox_polygon_spat)
plot(bbox_polygon_spat)



# DEM ==========================================================================
url <- "https://elevation-direct-downloads.s3-ap-southeast-2.amazonaws.com/1sec-dem/70715.zip"
# First accessed: 28th May 2025
# Geoscience Australia - SRTM-derived 1 Second Digital Elevation Models Version 1.0
# WGS 84 (EPSG:4326)
# DEM-S (smoothed dem)
# Approx. 30m res
# Potentially upsample to 10m resolution using bilenar to match high res 10m EVI layer

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(),"Bremer Bay spatial layers/DEM/dem")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "srtm-1sec-dems-v1-COG.zip")
dest_file

# Download the file
download.file(url, destfile = dest_file, mode = "wb")

# Unzip the file
unzip(dest_file, exdir = dest_dir)

# Optional: Remove the zip file after extraction
file.remove(dest_file)

# Identify the .tif file (assuming only one .tif in folder)
tif_file <- list.files(dest_dir, pattern = "^srtm-1sec-dems-v1-COG.tif$", full.names = TRUE)
tif_file

# Read raster DEM
dem <- rast(tif_file)

# Check original CRS
st_crs(dem)  # Should be UTM zone 50 if relevant

# Plot
plot(dem)

# Define large cropping extent as an sf polygon
large_crop_box <- st_as_sfc(st_bbox(c(
  xmin = 115,
  xmax = 120,
  ymin = -31,
  ymax = -36
), crs = 4326))

# Create crop vector
crop_vect <- vect(large_crop_box)

# Crop DEM
dem_cropped <- crop(dem, crop_vect)

# Plot
plot(dem_cropped)

# Save output raster
writeRaster(dem_cropped, filename = file.path(dest_dir, "SW_WA_DEM_30m_WGS84_EPSG4326.tif"), overwrite = TRUE)


# Crop DEM for Bremer Bay
# Define study area cropping extent as an sf polygon
# Transform crop extent to match DEM CRS (UTM zone 50)
bbox_4326 <- st_transform(bbox_polygon, crs(dem))
crop_vect_4326 <- vect(bbox_4326)

# Crop DEM
dem_cropped_4326 <- crop(dem, crop_vect_4326) 

# Plot
plot(dem_cropped_4326)

# Project raster
dem_cropped <- project(dem_cropped_4326, "EPSG:7850", method = "bilinear", res = 30)

# Mask
dem_cropped <- crop(dem_cropped, bb_study_area_7850)
dem_cropped <- mask(dem_cropped, bb_study_area_7850)

# Plot
plot(dem_cropped)

# Save output raster
writeRaster(dem_cropped, filename = file.path(dest_dir, "Bremer_Bay_DEM_30m_GDA2020_UTM_zone_50_EPSG7850.tif"), overwrite = TRUE)

dem_cropped <- rast(file.path(getwd(), "Bremer Bay spatial layers/DEM/dem/Bremer_Bay_DEM_30m_GDA2020_UTM_zone_50_EPSG7850.tif"))




# Upsampled DEM to 10m res - needs to be projected anyway so can upsample projected dem
# Original DEM in EPSG:7850 with 30 m resolution

# Create a 10 m resolution template using `ext()` not `extent()`
template <- rast(ext(dem_cropped))  # get extent
res(template) <- 10                 # set 10 m resolution
crs(template) <- crs(dem_cropped)  # match CRS

# Resample using bilinear interpolation
dem_10m <- resample(dem_cropped, template, method = "bilinear")

# Plot
plot(dem_10m, main = "10 m DEM (Resampled from 30 m using bilinear interpolation)")
dem_10m


# # DEM 2 ========================================================================
# # DEM from 10m interpolated contours based on 10m grid
# # https://catalogue.data.wa.gov.au/dataset/dpird-10-metre-contours
# 
# # Load contour shapefile
# contours <- st_read("F:\\lemerson\\R\\Eagles and Pumas\\Bremer Bay spatial layers\\Downloaded zip files\\Contours_10_m_DPIRD_073_WA_GDA2020_Public_Shapefile (2)\\Contours_10_m_DPIRD_073.shp")  # must have elevation attribute
# 
# # Reproject and crop
# contours_proj <- st_transform(contours, crs = st_crs(bb_study_area_7850))
# contours_cropped <- st_intersection(contours_proj, bb_study_area_7850)
# 
# # Filter LINESTRING only
# contours_lines <- contours_cropped[st_geometry_type(contours_cropped) == "LINESTRING", ]
# 
# # Sample points every ~10 m
# points_sampled <- st_line_sample(st_geometry(contours_lines), density = 1 / 10)
# 
# # Check number of samples: if too many, consider density = 1/20 or 1/30
# lengths(points_sampled)
# 
# # Create sfc of all sampled points
# points_geom <- do.call(c, points_sampled)
# points_geom <- st_sfc(points_geom, crs = st_crs(contours_lines))
# 
# # Match elevation to points
# counts <- lengths(points_sampled)
# elevations <- rep(contours_lines$elevation, counts)
# 
# # sf object
# elev_points_sf <- st_sf(elevation = elevations, geometry = points_geom)
# 
# # Convert to terra object
# elev_points_vect <- vect(elev_points_sf) # Falls over here - takes forever and may not even run at all!!
# 
# # Create empty raster template at 10 m resolution
# dem_raster <- rast(ext(elev_points_vect), resolution = 10, crs = crs(elev_points_vect))
# 
# # Interpolate using terra's IDW
# dem_from_contours <- interpolate(dem_raster, elev_points_vect, method = "idw", field = "elevation")
# 
# # Save
# writeRaster(dem_from_contours, "contour_based_dem_10m.tif", overwrite = TRUE)





# Compare ==

# Elvation differences
# Ensure both rasters have same extent/resolution
dem_upsampled <- resample(dem_upsampled, dem_contour)

# Difference raster
dem_diff <- dem_contour - dem_upsampled
plot(dem_diff)

# Statistics
summary(values(dem_diff), na.rm = TRUE)

# Profile analysis
line <- drawLine()  # interactively draw a line in `terra`

profile_contour <- extract(dem_contour, line)
profile_upsampled <- extract(dem_upsampled, line)

plot(profile_contour[,2], type="l", col="blue", ylab="Elevation (m)", xlab="Point along line")
lines(profile_upsampled[,2], col="red")
legend("topright", legend=c("Contour DEM", "Upsampled DEM"), col=c("blue", "red"), lty=1)


# Slope and curvature comparison
slope_contour <- terrain(dem_contour, v="slope", unit="degrees")
slope_upsampled <- terrain(dem_upsampled, v="slope", unit="degrees")

curvature_contour <- terrain(dem_contour, v="curvature")
curvature_upsampled <- terrain(dem_upsampled, v="curvature")

# Compare histograms or differences
hist(values(slope_contour), breaks=100, main="Contour Slope", col=adjustcolor("blue", 0.5))
hist(values(slope_upsampled), breaks=100, add=TRUE, col=adjustcolor("red", 0.5))




# Calculate TRI, TPI and Slope, SD_slope =======================================

study_location <- "Bremer Bay"

# Create slope raster -------------------------------------------------------
# Define name of folder for saving
data_type <- "Slope"

# Generate raster
spatial_variable <- terra::terrain(dem_cropped, v="slope", unit="degrees", neighbors=8)
# slope_variable <- raster::terrain(dsm, opt="slope", unit="degrees", neighbors=8)

# Checks
spatial_variable
class(spatial_variable)
plot(spatial_variable)
summary(values(spatial_variable))
global(spatial_variable, fun="isNA")
# Crop and mask - should not change because already using cropped and masked dem
cropped <- terra::crop(spatial_variable, bb_study_area_7850)
masked <- terra::mask(cropped, bb_study_area_7850)
plot(masked)
masked
summary(values(masked))
global(masked, fun="isNA")

# Save raster to file 
# Construct the file path
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)

# Create the folder if it doesn't exist
subfolder_path <- file.path(folder_path, data_type)
dir.create(subfolder_path, recursive = TRUE, showWarnings = FALSE)

# Save the object to the specified file path
writeRaster(masked, file_path, overwrite = T)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Slope"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
slope <- terra::rast(file_path)




# Create SD of slope raster --------------------------------------------------
# Define variable of interest
data_type <- "SD slope" # SD of slope

# Generate raster
spatial_variable <- terra::focal(slope, w = matrix(1, nrow = 5, ncol = 5), fun = sd, na.rm = TRUE)

# Checks
spatial_variable
class(spatial_variable)
plot(spatial_variable)
summary(values(spatial_variable))
global(spatial_variable, fun="isNA")
# Crop and mask - should not change because already using cropped and masked dem - althugh NAs might change
cropped <- terra::crop(spatial_variable, bb_study_area_7850)
masked <- terra::mask(cropped, bb_study_area_7850)
plot(masked)
masked
summary(values(masked))
global(masked, fun="isNA")

# Save raster to file 
# Construct the file path
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)

# Create the folder if it doesn't exist
subfolder_path <- file.path(folder_path, data_type)
dir.create(subfolder_path, recursive = TRUE, showWarnings = FALSE)

# Save the object to the specified file path
writeRaster(masked, file_path, overwrite = T)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "SD Slope"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
SD_slope <- terra::rast(file_path)




# Create topographic position index raster ---------------------------------------------------
# Define name of folder for saving
data_type <- "TPI"

# Generate raster
spatial_variable <- terra::terrain(dem_cropped, v="TPI") # Undefined moving window size - i think it might be default of 3 x 3...?
# spatial_variable <- raster::focal(dsm, w = matrix(1, nrow = 5, ncol = 5), fun = function(x) x[13] - mean(x[-13])) # This is 5 x 5 cells with central cell being position 13

# Checks
spatial_variable
class(spatial_variable)
plot(spatial_variable)
summary(values(spatial_variable))
global(spatial_variable, fun="isNA")
# Crop and mask - should not change because already using cropped and masked dem - althugh NAs might change
cropped <- terra::crop(spatial_variable, bb_study_area_7850)
masked <- terra::mask(cropped, bb_study_area_7850)
plot(masked)
masked
summary(values(masked))
global(masked, fun="isNA")

# Save raster to file 
# Construct the file path
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)

# Create the folder if it doesn't exist
subfolder_path <- file.path(folder_path, data_type)
dir.create(subfolder_path, recursive = TRUE, showWarnings = FALSE)

# Save the object to the specified file path
writeRaster(masked, file_path, overwrite = T)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "TPI"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
TPI <- terra::rast(file_path)


# Create terrain ruggedness index raster ---------------------------------------------------
# Define name of folder for saving
data_type <- "TRI"

# Generate raster
spatial_variable <- terra::terrain(dem_cropped, v="TRI") # Undefined moving window size - i think it might be default of 3 x 3...?
# spatial_variable <- raster::focal(dsm, w = matrix(1, nrow = 5, ncol = 5), fun = function(x) x[13] - mean(x[-13])) # This is 5 x 5 cells with central cell being position 13

# Checks
spatial_variable
class(spatial_variable)
plot(spatial_variable)
summary(values(spatial_variable))
global(spatial_variable, fun="isNA")
# Crop and mask - should not change because already using cropped and masked dem - althugh NAs might change
cropped <- terra::crop(spatial_variable, bb_study_area_7850)
masked <- terra::mask(cropped, bb_study_area_7850)
plot(masked)
masked
summary(values(masked))
global(masked, fun="isNA")

# Save raster to file 
# Construct the file path
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)

# Create the folder if it doesn't exist
subfolder_path <- file.path(folder_path, data_type)
dir.create(subfolder_path, recursive = TRUE, showWarnings = FALSE)

# Save the object to the specified file path
writeRaster(masked, file_path, overwrite = T)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "TRI"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
TRI <- terra::rast(file_path)






# Fire history =================================================================
url <- "https://direct-download.slip.wa.gov.au/datadownload/Environment/DBCA_Fire_History_DBCA_060_WA_GDA2020_Public_Shapefile.zip"
# First accessed: 28th May 2025
# DataWA - DBCA Fire History (DBCA-060)
# GDA2020 (EPSG:7844)
# Data last updated: 28-01-2025

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/Fire")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "DBCA_Fire_History_DBCA_060.shp")

# Download the file
download.file(url, destfile = dest_file, mode = "wb")

# Unzip the file
unzip(dest_file, exdir = dest_dir)

# Optional: Remove the zip file after extraction
file.remove(dest_file)

# Read the shapefile
fire <- st_read(dest_file)

# Check the CRS
st_crs(fire)

# Structure
str(fire)
class(fire)

# Transform crop extent to match DEM CRS (UTM zone 50)
fire_proj <- st_transform(fire, crs = st_crs(bb_study_area_7850))

# # Fix geometries
# fire_proj <- st_make_valid(fire_proj)

# Crop
st_crs(fire_proj)==st_crs(bb_study_area_7850) # Check geometries match
fire_cropped <- st_intersection(fire_proj, bb_study_area_7850)

# Column names
colnames(fire_cropped)

# Define your reference date (January 28, 2025)
reference_date <- as.Date("2025-01-28")

# Convert fih_date1 to Date format if it is not already
fire_cropped$fih_date1 <- as.Date(fire_cropped$fih_date1)

# Calculate time since fire in years
fire_cropped <- fire_cropped %>%
  mutate(time_since_fire_yrs = as.numeric(difftime(reference_date, fih_date1, units = "days")) / 365.25)

# Check
fire_cropped %>%
  select(fih_date1, time_since_fire_yrs) %>%
  head()

fire_cropped$time_since_fire_yrs

# Remove other columns
fire_cropped <- fire_cropped %>%
  select(time_since_fire_yrs)

# Plot
head(fire_cropped)
plot(fire_cropped, main = "Time since fire (Years)")

# Save output 
st_write(fire_cropped,
         dsn = file.path(dest_dir, "Bremer Bay fire.gpkg"),
         layer = "time_since_fire_cropped_7850",
         driver = "GPKG",
         delete_layer = TRUE)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Fire"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".gpkg", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
TSF <- st_read(file_path, layer = "time_since_fire_cropped_7850")
plot(TSF)
class(TSF)


# Now convert to terra
TSF_spat <- vect(TSF_sf)
class(TSF_spat)
str(TSF_spat)
plot(TSF_spat)
plot(TSF_spat, "time_since_fire_yrs", type = "continuous", col = terrain.colors(44))



# Check layers are aligning
tm_shape(dem_cropped) + 
  tm_raster(title = "DEM Elevation") +
  tm_shape(bb_study_area_7850) + 
  tm_borders(col = "black", lwd = 2) +
  tm_shape(fire_cropped) + 
  tm_borders(col = "red", lwd = 2)



# Habitat types defined as either native vegetation, or farmland 
# and plantation using the Native Vegetation Extent layer for Western Australia 
# (DPIRD 005) and landuse categories from ABARES (2022). For non-native vegetation 
# broadscale landuse data were used to assign categories of cropping, plantation 
# or grazing (ABARES, 2022). Rehabilitation sites within the Fitz-Stirling were 
# mapped using data from ABARES and validated using orthophotos.


# Native Vegetation extent =====================================================
# url <- "https://catalogue.data.wa.gov.au/dataset/native-vegetation-extent"
# First accessed: 28th May 2025
# DataWA - Native Vegetation Extent (DPIRD-005)
# Department of Primary Industries and Regional Development
# GCS_GDA_2020

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/Native veg")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "Native_Veg_Extent_DPIRD_005.shp")
dest_file

# # Download the file
# download.file(url, destfile = dest_file, mode = "wb")
# 
# # Unzip the file
# unzip(dest_file, exdir = dest_dir)
# 
# # Optional: Remove the zip file after extraction
# file.remove(dest_file)

# Read the shapefile
native_veg <- st_read(dest_file)

# Check the CRS
st_crs(native_veg)

# Structure
str(native_veg)

# Transform to match DEM CRS (UTM zone 50)
native_veg <- st_transform(native_veg, crs = st_crs(bb_study_area_7850))
st_crs(native_veg)

# # Fix geometries
# native_veg <- st_make_valid(native_veg)

# # Save output 
# st_write(native_veg,
#          dsn = file.path(dest_dir, "native_veg.gpkg"),
#          layer = "native_veg",
#          driver = "GPKG",
#          delete_layer = TRUE)

# Crop
st_crs(native_veg)==st_crs(bb_study_area_7850) # Check geometries match
native_veg_cropped <- st_intersection(native_veg, bb_study_area_7850)
class(native_veg_cropped)

# Plot
colnames(native_veg_cropped)
plot(native_veg_cropped["comments"], main = "Native vegetation extent")

# Save output raster
st_write(native_veg_cropped,
         dsn = file.path(dest_dir, "native_veg_cropped.gpkg"),
         layer = "native_veg_cropped",
         driver = "GPKG",
         delete_layer = TRUE)

# Convert sf to SpatVector (terra format)
native_veg_cropped <- vect(native_veg_cropped)

# Rasterize vector layer and calculate proportional values: binary values (1 = native veg, 0 = non-veg)
# Rasterize with 'dem_cropped' as the template
# The default rasterization will assign NA or 1 depending on presence of polygons
# But terra sometimes expects a field - so you can create a dummy field:
native_veg_cropped$burn <- 1
native_veg_raster <- terra::rasterize(native_veg_cropped, dem_cropped, field="burn", background=0)

# Crop and mask
native_veg_raster <- terra::crop(native_veg_raster, bb_study_area_7850)
native_veg_raster <- terra::mask(native_veg_raster, bb_study_area_7850)

# Plot to check
plot(native_veg_raster, main = "Native Vegetation (Binary Raster)")

# Convert
native_veg_raster <- raster::raster(native_veg_raster)

# Create buffer weight matrix (100 m radius)
buffer <- raster::focalWeight(native_veg_raster, 100, "circle")
dim(buffer)  # How many rows and cols?
sum(buffer)  # Should be close to 1 if normalized

# Calculate proportional cover
native_veg_prop <- raster::focal(native_veg_raster, w=buffer, na.rm=TRUE, pad=TRUE, padValue=0)

# Plot them using simple raster algebra to check that no cell sums to greater than 1
plot(native_veg_prop)

# Convert RasterLayer to SpatRaster
native_veg_prop <- rast(native_veg_prop)

# Crop and mask
native_veg_prop <- terra::crop(native_veg_prop, bb_study_area_7850)
native_veg_prop <- terra::mask(native_veg_prop, bb_study_area_7850)

# Plot
plot(native_veg_prop)

# Save
writeRaster(native_veg_prop,
            filename = file.path(dest_dir, "Bremer Bay native veg.tif"),
            filetype = "GTiff",
            overwrite = TRUE)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Native veg"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
native_veg_prop <- rast(file_path)
plot(native_veg_prop)
class(native_veg_prop)



# Roads ========================================================================
# url <- "https://digital.atlas.gov.au/datasets/digitalatlas::national-roads-2/about"
# First accessed: 30th May 2025
# Digital Atlas of Australia - National Roads
# Geoscience Australia 2021 
# Appears to be most comprehensive roads layer available but not all roads and trails included - 
# Need to add roads and tracks manually in QGIS depending on movement data from cats 
# and if roads appear relevant for analysis - may not be strongly selected for in these open landscapes.

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/Roads")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, "National_Roads_4244405608125821245.gpkg")

# Read the shapefile
roads <- st_read(dest_file)

# Check the CRS
st_crs(roads)

# Structure
str(roads)
class(roads)

# Transform to match DEM CRS (UTM zone 50)
roads_proj <- st_transform(roads, crs = st_crs(bb_study_area_7850))

# Crop
st_crs(roads_proj)==st_crs(bb_study_area_7850) # Check geometries match
roads_cropped <- st_intersection(roads_proj, bb_study_area_7850)

# Column names
colnames(roads_cropped)

# Remove other columns
# roads_cropped <- fire_cropped %>%
#   select(???)

# Plot
head(roads_cropped)
plot(roads_cropped["date_created"], main = "Roads")
class(roads_cropped)

# Save output 
st_write(roads_cropped,
         dsn = file.path(dest_dir, "Bremer Bay roads.gpkg"),
         layer = "roads_cropped_7850",
         driver = "GPKG",
         delete_layer = TRUE)


# Create proportion raster for roads
# Convert to SpatVector
roads_vect <- vect(roads_cropped)

# Buffer
roads_buffered <- buffer(roads_vect, width = 30)  
# width is radius, so 30 m each side - This width required to achieve connected pixels. Overestimates road width and coverage

# Check output
plot(roads_buffered)

# Add dummy field for rasterization
roads_buffered$road_cover <- 1

# Rasterize with DEM as template
roads_raster <- rasterize(roads_buffered, dem_cropped, field = "road_cover", background = 0)

# Crop and mask to study area
roads_raster <- crop(roads_raster, bb_study_area_7850)
roads_raster <- mask(roads_raster, bb_study_area_7850)

# Plot
plot(roads_raster, main = "Binary Road Raster (1 = road, 0 = no road)")

# Convert to raster format if needed
roads_raster <- raster::raster(roads_raster)

# Create 30 m circular weight matrix
w <- raster::focalWeight(roads_raster, 30, type = "circle") # Very small circle - 
# essentially saying whether or not ehy are using road - almost binary predictor. Could increase

# Calculate focal (moving window) average: gives proportion of road cover
road_prop_cover <- raster::focal(roads_raster, w = w, na.rm = TRUE, pad = TRUE, padValue = 0)

# Plot
plot(road_prop_cover, main = "Proportional Road Cover (30 m radius)")

# Convert back to SpatRaster
road_prop_cover <- rast(road_prop_cover)

# Final crop and mask (optional)
road_prop_cover <- crop(road_prop_cover, bb_study_area_7850)
road_prop_cover <- mask(road_prop_cover, bb_study_area_7850)

# Plot
plot(road_prop_cover, main = "Proportional Road Cover (30 m radius)")

# Save
writeRaster(road_prop_cover,
            filename = file.path(dest_dir, "Bremer Bay road cover.tif"),
            filetype = "GTiff",
            overwrite = TRUE)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Roads"
layer <- "road cover"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", layer, ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
road_cover <- terra::rast(file_path)
plot(road_cover)
class(road_cover)



# Create distance to road layer

# Optional: Buffer roads if you want to include a buffer zone around the roads (e.g., 15 meters)
# roads_vect <- buffer(roads_vect, width = 15)

# Create a template raster from your DEM (or any raster with matching extent and resolution)
r_template <- rast(dem_cropped)

# Rasterize roads: 1 for roads, NA for no roads
roads_raster_binary <- rasterize(roads_vect, r_template, field = 1, background = NA)

# Calculate distance to nearest road (cells with value 1)
distance_to_roads <- distance(roads_raster_binary)

# Optional - Crop and mask the distance raster to your study area
distance_to_roads <- crop(distance_to_roads, bb_study_area_7850)
distance_to_roads <- mask(distance_to_roads, bb_study_area_7850)

# Plot to check results
plot(distance_to_roads, main = "Distance to Roads")

# Save the distance raster
writeRaster(distance_to_roads,
            filename = file.path(dest_dir, "Bremer Bay distance to roads.tif"),
            filetype = "GTiff",
            overwrite = TRUE)

# Read in layer
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Roads"
layer <- "distance to roads"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", layer, ".tif", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
dist_road <- terra::rast(file_path)
plot(dist_road)
class(dist_road)




# Distance to freshwater body, wetland, riparian zone ==========================



# Calculating drainage lines from DEM ==========================================
# Install and load the whitebox package
# install.packages("whitebox")
# whitebox::install_whitebox()
library(whitebox)

# Set working directory (adjust path to where your DEM is stored)
wbt_set_working_dir(file.path(getwd(), "Bremer Bay spatial layers", "DEM", "dem"))
dir_path <- file.path(getwd(), "Bremer Bay spatial layers", "DEM", "dem")

# Set filenames
dem_file <- file.path(dir_path, "Bremer_Bay_DEM_30m_GDA2020_UTM_zone_50_EPSG7850.tif")
dem_filled <- file.path(dir_path, "Bremer_Bay_DEM_filled.tif")
flow_dir <- file.path(dir_path, "Bremer_Bay_flow_dir.tif")
flow_acc <- file.path(dir_path, "Bremer_Bay_flow_acc.tif")
streams_raster <- file.path(dir_path, "Bremer_Bay_streams.tif")
streams_vector <- file.path(dir_path, "Bremer_Bay_streams_vector.shp")
depressions_raster <- file.path(dir_path, "Bremer_Bay_depressions.tif")
depressions_vector <- file.path(dir_path, "Bremer_Bay_depressions_vector.shp")


# Step 1: Fill depressions
wbt_fill_depressions(
  dem = dem_file,
  output = dem_filled
)

# Read the filled DEM raster
dem_filled_raster <- rast(dem_filled)

# Plot
plot(dem_filled_raster, main = "Filled DEM")


# Step 2: Calculate flow direction (D8)
wbt_d8_pointer(
  dem = dem_filled,
  output = flow_dir
)

# Read the filled DEM raster
flow_dir_raster <- rast(flow_dir)

# Plot
plot(flow_dir_raster, main = "Flow")


# Step 3: Calculate flow accumulation
wbt_d8_flow_accumulation(
  input = dem_filled,
  output = flow_acc,
  out_type = "cells"
)

# Step 4: Extract stream network from flow accumulation
# Adjust threshold (e.g., 100) depending on your DEM resolution and landscape
wbt_extract_streams(
  flow_accum = flow_acc,
  output = streams_raster,
  threshold = 100
)

# Example of testing three thresholds
thresholds <- c(100, 250, 500, 750, 1000)

for (t in thresholds) {
  stream_output <- paste0("stream_network_", t, ".tif")
  wbt_extract_streams(
    flow_accum = flow_acc,
    output = file.path(dir_path, stream_output),
    threshold = t
  )
}


# Step 5: Identify depressions (potential water-holding areas)
wbt_depression_mapping(
  dem = dem_file,
  output = depressions_raster
)

# Optional: Convert streams to vector (shapefile)
wbt_raster_streams_to_vector(
  streams = streams_raster,
  d8_pntr = flow_dir,
  output = streams_vector
)

# Optional: Convert depression raster to polygons
wbt_raster_to_vector_polygons(
  input = depressions_raster,
  output = depressions_vector
)




# Enhanced Vegetation Index ====================================================
##  See more than 300+ examples of rgee at https://r-earthengine.github.io/.!!!
# Python environment setup (requires Python >= v3.5 ; will install Miniconda)
rgee::ee_clean_pyenv() # Do this first then restart R

# After running above and restarting R
rgee::ee_install() # Then this and restart

# Downgrade ee api
reticulate::py_install('earthengine-api==0.1.370', envname = 'rgee', method = 'conda')
rgee::ee_check()


### Try starting here but if there are issues re-install python ###
# After running above and restarting - Ensure reticulate is using the correct conda binary - i.e. correct version of python
reticulate::use_condaenv("rgee", required = TRUE)
reticulate::py_config()  # Verify it's using the correct Python
rgee::ee_check()

# Remove credentials before updating gcloud
rgee::ee_clean_user_credentials('lukedanielemerson1982@gmail.com') #only fist time or need to clear credentials
rgee::ee_Authenticate(
  scopes = c(
    "https://www.googleapis.com/auth/drive",
    "https://www.googleapis.com/auth/earthengine"
  )
)

# Initialize Earth Engine and GD :  /!\/!\/!\ NEED TO RUN RSTUDIO IN ADMINISTRATOR MODE FOR THIS LINE /!\/!\/!\
library(rgee)
rgee::ee_Initialize(user = "lukedanielemerson1982@gmail.com", drive = TRUE, gcs = FALSE)

# Get your Earth Engine user info to confirm
ee_user_info()

# Check users, credentials, path 
rgee::ee_users()
rgee::ee_get_earthengine_path()
rgee::ee_check()
rgee::ee_check_python()
rgee::ee_check_credentials()
rgee::ee_check_python_packages()

# Checking installation non-R dependencies
reticulate::conda_list()
reticulate::py_available()
reticulate::py_config()
reticulate::py_discover_config()
reticulate::py_config()$version

# Load libraries
library(rgee)
library(sf)
library(dplyr)
library(rgee)
library(raster)
library(terra)
library(sf)
library(dplyr)
library(rgeeExtra) # Not available for this version of R
library(tidyrgee)
library(googleCloudStorageR)
library(rgee)
library(googledrive)
library(reticulate)
library(tidyverse)
library(here)
library(geojsonio)
library(devtools)

# Test code to check things are working
srtm <- rgee::ee$Image("USGS/SRTMGL1_003")

viz <- list(
  max = 4000,
  min = 0,
  palette = c("#000000","#5AAD5A","#A9AD84","#FFFFFF")
)

rgee::Map$addLayer(
  eeObject = srtm,
  visParams =  viz,
  name = 'SRTM',
  # legend = TRUE
)


# Read in layer 
# Construct the file path
study_location <- "Bremer Bay"
data_type <- "Study area bounding box"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".shp", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)
file_path

# Read file and transform CRS to WGS84 to align with gee
aoi_sf <- sf::st_read(file_path) %>%
  st_transform(crs = 4326)


# Check crs
sf::st_crs(aoi_sf)

# Convert to gee object
aoi <- sf_as_ee(aoi_sf)
class(aoi)

# Visualise and ensure its in correct location
Map$centerObject(aoi)
Map$addLayer(aoi)

# Define date parameters (can be modified later)
start_date <- '2025-01-01'
end_date <- '2025-06-06'  # change this to '2024-07-01' for mid last year to Oct this year

# # Alternative dates once data has been collected across sites
# start_date <- '2024-07-01'
# end_date <- '2024-10-31'


# Load Sentinel-2 surface reflectance
s2_sr <- ee$ImageCollection("COPERNICUS/S2_SR")$
  filterBounds(aoi)$
  filterDate(start_date, end_date)$
  filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
  map(function(img) {
    # Scale reflectance bands
    img <- img$divide(10000)
    
    # Calculate EVI
    evi <- img$expression(
      "2.5 * ((NIR - RED) / (NIR + 6 * RED - 7.5 * BLUE + 1))",
      list(
        NIR = img$select("B8"),
        RED = img$select("B4"),
        BLUE = img$select("B2")
      )
    )$rename("EVI")
    
    img$addBands(evi)
  })

# Create a median composite of EVI
evi_median <- s2_sr$select("EVI")$median()$clip(aoi)

# Visualize in RStudio viewer
Map$centerObject(aoi)
Map$addLayer(evi_median,
             visParams = list(min = 0, max = 1, palette = c("white", "green")),
             name = "Median EVI")

# Get the number of images in the filtered ImageCollection used for EVI calculation
s2_sr$size()$getInfo()

# Export to Google Drive (optional)
task <- ee_image_to_drive(
  image = evi_median,
  description = "median_EVI_2025",
  folder = "rgee_evi",
  fileNamePrefix = "EVI_2025_median",
  region = aoi$geometry(),
  scale = 10,
  maxPixels = 1e13
)
task$start()

# Monitor the task
ee_monitoring(task)

# Access rgee files
# Gdrive access
googledrive::drive_user()
googledrive::drive_auth(email = "lukedanielemerson1982@gmail.com")

# Check folders in gdrive to check connection
rgee_folder <- googledrive::drive_get("rgee_evi")
files_in_rgee <- googledrive::drive_ls(path = rgee_folder)
print(files_in_rgee)

# Create folder path
target_name <- files_in_rgee$name
target_file <- googledrive::drive_ls(path = rgee_folder) %>%
  dplyr::filter(name == target_name)

# Define the destination directory (Data folder in current working directory)
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/EVI")
if (!dir.exists(dest_dir)) dir.create(dest_dir)

# Define the destination file path
dest_file <- file.path(dest_dir, target_name)

# Download it to your working directory
googledrive::drive_download(file = target_file, path = dest_file, overwrite = TRUE)

# Read raster
evi <- rast(dest_file)
evi
plot(evi)

summary(evi)

# Need to work out why some values are above 1. 
# Try extracting water areas
# Check what is highest evi areas
# Crop and mask

evi_water <- classify(evi, cbind(-Inf, 0, 1))  # water = 1
plot(evi_water)



# Satellite basemap imagery ====================================================
# Load required libraries
library(rgee)
library(sf)
library(googledrive)
library(dplyr)
library(raster)
library(terra)
library(rgeeExtra) # Not available for this version of R
library(tidyrgee)
library(googleCloudStorageR)
library(googledrive)
library(reticulate)
library(tidyverse)
library(here)
library(geojsonio)
library(devtools)

### Try starting here but if there are issues re-install python ###
# After running above and restarting - Ensure reticulate is using the correct conda binary - i.e. correct version of python
reticulate::use_condaenv("rgee", required = TRUE)
reticulate::py_config()  # Verify it's using the correct Python
rgee::ee_check()

# Remove credentials before updating gcloud
rgee::ee_clean_user_credentials('lukedanielemerson1982@gmail.com') #only fist time or need to clear credentials
rgee::ee_Authenticate(
  scopes = c(
    "https://www.googleapis.com/auth/drive",
    "https://www.googleapis.com/auth/earthengine"
  )
)

# Initialize Earth Engine and GD :  /!\/!\/!\ NEED TO RUN RSTUDIO IN ADMINISTRATOR MODE FOR THIS LINE /!\/!\/!\
rgee::ee_Initialize(user = "lukedanielemerson1982@gmail.com", drive = TRUE, gcs = FALSE)

# Get your Earth Engine user info to confirm
ee_user_info()

# Check users, credentials, path 
rgee::ee_users()
rgee::ee_get_earthengine_path()
rgee::ee_check()
rgee::ee_check_python()
rgee::ee_check_credentials()
rgee::ee_check_python_packages()

# Checking installation non-R dependencies
reticulate::conda_list()
reticulate::py_available()
reticulate::py_config()
reticulate::py_discover_config()
reticulate::py_config()$version

# Test code to check things are working
srtm <- rgee::ee$Image("USGS/SRTMGL1_003")

viz <- list(
  max = 4000,
  min = 0,
  palette = c("#000000","#5AAD5A","#A9AD84","#FFFFFF")
)

rgee::Map$addLayer(
  eeObject = srtm,
  visParams =  viz,
  name = 'SRTM',
  # legend = TRUE
)


# Define study area path and read shapefile
study_location <- "Bremer Bay"
data_type <- "Study area bounding box"
working_dir <- getwd()
folder_path <- file.path(paste(study_location, "spatial layers", sep = " "))
file_name <- paste(study_location, " ", tolower(data_type), ".shp", sep = "")
file_path <- file.path(working_dir, folder_path, data_type, file_name)

aoi_sf <- sf::st_read(file_path) %>%
  st_transform(crs = 4326)

# Check crs
sf::st_crs(aoi_sf)

# Convert to gee object
aoi <- sf_as_ee(aoi_sf)
class(aoi)

# Visualise and ensure its in correct location
Map$centerObject(aoi)
Map$addLayer(aoi)

# Define time range for June 2025
start_date <- '2025-05-01'
end_date <- '2025-06-30'

# Filter Sentinel-2 SR collection for RGB bands (true color)
s2_rgb <- ee$ImageCollection("COPERNICUS/S2_SR")$
  filterBounds(aoi)$
  filterDate(start_date, end_date)$
  filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 5))$
  map(function(img) {
    img$divide(10000)$select(c("B4", "B3", "B2"))$rename(c("Red", "Green", "Blue"))
  })

# Create median composite and clip to AOI
rgb_median <- s2_rgb$median()$clip(aoi)

# Visualize in R
Map$centerObject(aoi)
Map$addLayer(
  rgb_median,
  visParams = list(min = 0, max = 0.3, bands = c("Red", "Green", "Blue")),
  name = "Sentinel-2 RGB June 2025"
)

# Count number of images used
s2_rgb$size()$getInfo()  # Prints number of images used

# Export the image to Google Drive
task_rgb <- ee_image_to_drive(
  image = rgb_median,
  description = "RGB_May_June_2025",
  folder = "rgee_rgb",
  fileNamePrefix = "RGB_May_June_2025_median",
  region = aoi$geometry(),
  scale = 10,
  maxPixels = 1e13
)
task_rgb$start()

# Monitor task
ee_monitoring(task_rgb)

# OPTIONAL: After export completes, download it from Google Drive
# Get exported file name
rgee_folder <- googledrive::drive_get("rgee_rgb")
files_in_rgee <- googledrive::drive_ls(path = rgee_folder)
target_name <- files_in_rgee$name
target_file <- googledrive::drive_ls(path = rgee_folder) %>%
  dplyr::filter(name == target_name)

# Set local download path
dest_dir <- file.path(getwd(), "Bremer Bay spatial layers/RGB")
if (!dir.exists(dest_dir)) dir.create(dest_dir)
dest_file <- file.path(dest_dir, target_name)

# Download file to disk
googledrive::drive_download(file = target_file, path = dest_file, overwrite = TRUE)

# Load the raster
rgb_raster <- rast(dest_file)

# Check the number of bands and value range
print(rgb_raster)
summary(rgb_raster)

# Optional: If values are between 0–1, scale to 0–255 for better visualization
if (max(values(rgb_raster), na.rm = TRUE) <= 1.5) {
  rgb_raster <- rgb_raster * 255
}

# Plot with RGB coloring
plotRGB(rgb_raster, r = 1, g = 2, b = 3, scale = 255, stretch = "lin", main = "True Color RGB Composite")



# Extraction ===================================================================
library(sf)
library(dplyr)

# Get all intersecting polygons per point
intersections <- st_intersects(point_locations, fire_cropped)

# Extract and summarize (e.g., minimum time since fire)
tsf_by_point <- lapply(intersections, function(idxs) {
  if (length(idxs) == 0) return(NA)
  min(fire_cropped$time_since_fire_yrs[idxs], na.rm = TRUE)
})

# Convert to numeric vector or bind back to your point data
point_locations$min_time_since_fire <- unlist(tsf_by_point)




```


